\documentclass{article}
\usepackage[margin=1in]{geometry}
\usepackage{setspace}
\usepackage{amsmath}

\def\DD{{\rm DD}}
\def\nn{\nonumber}

\title{Notes on tree dedispersion}
\author{Kendrick Smith}

\begin{document}

\maketitle

\setstretch{1.05}   % spacing between lines
\setlength{\parskip}{2.5pt}  % spacing between paragraphs

\section{Tree gridding kernel}

Our starting point for tree dedispersion is a shape $(N_{\rm freq}, N_{\rm time})$ array representing channelized intensity time series. We do not assume that frequency channels are evenly spaced (in CHORD we use unequal spacing).
We want to search this array for FRBs with delay $\propto \nu^{-2}$.

The first step is a ``tree gridding'' kernel which changes variables so that the delay is linear (i.e.~the FRB dispersion sweep looks like a straight line).
The input of the tree gridding kernel is a shape $(N_{\rm freq}, N_{\rm time})$ array, and the output is a shape $(N_{\rm tree}, N_{\rm time})$ array, where $N_{\rm tree}=2^r$.
The parameter $r$ is the tree dedispersion ``rank'' and is specified in a config file.

The leading axis of the input array represents frequency channels, and will be indexed by $0 \le F < N_{\rm freq}$.
The leading axis of the output array represents a reparameterized frequency channel (``tree-freq'' channel for short), and will be indexed by $0\le f < 2^r$.
Conceptually, the tree gridding kernel just reparameterizes the frequency axis (i.e.\ the time axis is a spectator.)

The tree gridding kernel also takes an auxiliary array \verb|channel_map|, which describes the mapping between frequency channels $0 \le F < N_{\rm freq}$ and tree-freq channels $0 \le f < N_{\rm tree}$.
The array \verb|channel_map| has length $N_{\rm tree}+1$, and its entries are (in general non-integer) frequency-channel indices in the range $[0, N_{\rm freq}]$.
Writing $c_f$ for the $f$-th entry, tree-freq channel $f$ is defined to occupy the half-open interval of frequency channels
\begin{equation}
I_f = \big[\, c_{f+1},\ c_f \,\big), \qquad 0 \le f < N_{\rm tree},
\end{equation}
i.e.~consecutive entries give the lower and upper edges of each tree-freq channel.
The entries are fixed by the dispersion geometry: the tree-freq channels are equally spaced in the dispersion variable $x \equiv \nu^{-2}$ (this is the change of variables that linearizes the delay), so the $f$-th edge sits at
\begin{equation}
x_f = \nu_{\rm hi}^{-2} + \frac{f}{N_{\rm tree}}\left(\nu_{\rm lo}^{-2} - \nu_{\rm hi}^{-2}\right),
\end{equation}
and $c_f$ is the frequency-channel index of the frequency $\nu_f = x_f^{-1/2}$, where the index increases with frequency, running from $0$ at the band bottom $\nu_{\rm lo}$ to $N_{\rm freq}$ at the band top $\nu_{\rm hi}$.
In particular $c_0 = N_{\rm freq}$ and $c_{N_{\rm tree}} = 0$, so \verb|channel_map| is monotonically {\em decreasing} ($c_{f+1} < c_f$).
We store it in double precision, since the gridding weights below are obtained by differencing adjacent entries, which loses relative precision.

The details of the tree gridding kernel are as follows.
Each output tree-freq channel $f$ is a weighted sum of the input frequency channels whose bins overlap the interval $I_f = [c_{f+1}, c_f)$:
\begin{equation}
{\rm out}[f] = \sum_{F=0}^{N_{\rm freq}-1} w_{fF}\, {\rm in}[F],
\end{equation}
where ${\rm in}[F]$ and ${\rm out}[f]$ are length-$N_{\rm time}$ time series, and the same weights are applied at every time sample (the time axis is untouched).
The weight $w_{fF}$ is the length of the overlap between the tree-freq channel interval $I_f$ and the unit-width frequency bin $[F, F+1)$:
\begin{equation}
w_{fF} = \max\!\big(\,0,\ \min(c_f,\, F+1) - \max(c_{f+1},\, F)\,\big).
\end{equation}
A frequency channel lying entirely within $I_f$ contributes with weight $1$, while a channel straddling an edge of $I_f$ contributes a fractional weight.
Because the intervals $I_f$ tile $[0, N_{\rm freq}]$, the weights of any fixed input channel sum to one, $\sum_f w_{fF} = 1$; the gridding kernel is thus a delay-linearizing rebinning of the frequency axis that conserves total intensity, $\sum_f {\rm out}[f] = \sum_F {\rm in}[F]$.

Note that the tree gridding kernel is a monotone {\em decreasing} reindexing operation.
Thus, to search for FRBs with positive dispersion (lowest frequencies arise last), we want to search over lines in the tree array with positive slope (highest tree-freq channels arrive last).

\section{Tree dedispersion}
\label{sec:tree_dedispersion}

After the tree gridding kernel, the next step is ``tree dedispersion'', which searches for straight lines with $0\le \mbox{slope} \le 1$.
(The extension to slope $>1$ will be discussed in a future section, along with more details such as subband searches, early triggers, and peak-finding.)

The dedispersion input array has shape $(2^r, N_{\rm time})$, where the leading axis is ``tree-freq'' (see above).
Straight lines will be parameterized by an integer $0 \le d < 2^r$, representing
the delay (in time samples) between the lowest and highest tree-freq channel,
and a time index $t$ representing the arrival time in the {\em highest} tree-freq channel
(i.e. the {\em latest} arrival time).
Thus, the dedispersion output array also has shape $(2^r, N_{\rm time})$, where the first axis represents a delay (or DM).\footnote{Frequently in the code (but not in these notes), it's convenient to use a {\em bit-reversed} delay index $0 \le d_{\rm rev} < 2^r$. (This turns out to allow some array operations to operate ``in place'', facilitating parallelization.)}

Tree dedispersion is a composition of functions:
\begin{equation}
\big( \mbox{Input array} \big)
  \overset{\DD(0)}{\longrightarrow} \big( \mbox{Intermediate array} \big)
  \overset{\DD(1)}{\longrightarrow} \cdots 
  \overset{\DD(r-1)}{\longrightarrow}
\big( \mbox{Output array} \big)
\label{eq:DD_composition}
\end{equation}
where the function $\DD(k)$ is described as follows:

\begin{itemize}

\item
The $\DD(k)$ input array is a shape $(2^{r-k}, 2^k, N_{\rm time})$ array.
The first axis represents a coarse frequency channel.
Each coarse-freq channel $f$ corresponds to a block of $2^k$ tree-freq channels
in the original tree input array:
\begin{equation}
2^k \, f \le (\mbox{tree-freq index}) < 2^k (f+1)  \label{eq:coarse_freq}
\end{equation}
The second axis represents a delay $0 \le d < 2^k$ within the coarse-freq channel
(\ref{eq:coarse_freq}). As $k$ increases, the coarse-freq channels double in width,
and we need twice as many delay indices to cover the range $0 \le (\mbox{slope}) \le 1$.

The time index represents an arrival time at the largest tree-freq channel
in the coarse-freq channel (\ref{eq:coarse_freq}), i.e.~the latest arrival time within the coarse-freq channel.

\item The $\DD(k)$ operation is defined by:

\begin{align}
{\tt out}[f,2d,t] &= \frac{1}{\sqrt{2}} 
  \big( {\tt in}[2f,d,t-d] + {\tt in}[2f+1,d,t] \big) \nn \\
{\tt out}[f,2d+1,t] &= \frac{1}{\sqrt{2}} 
  \big( {\tt in}[2f,d,t-d-1] + {\tt in}[2f+1,d,t] \big)
\end{align}

\item
As formulated above, tree dedispersion can lead to negative time indices.
By default, \verb|in[...,-N]| is defined to be zero.
(In the real-time system, we run tree dedispersion in an ``incremental'' context where a negative time index should be interpreted as a reference to data received in a previous time chunk.)

\item The $\DD(0)$ input array is sometimes reshaped from $(2^r,1,N_{\rm time})$
 to $(2^r,N_{\rm time})$, and the $\DD(r-1)$ output array is sometimes reshaped
 from $(1,2^r,N_{\rm time})$ to $(2^r,N_{\rm time})$.
 Thus, we think of the entire tree dedispersion operation (\ref{eq:DD_composition}) as having input and output shapes $(2^r, N_{\rm time})$.

\end{itemize}

\section{Subband searches}

So far, we have described a search for FRBs which cover the full tree-freq range $0 \le f < 2^r$.
This would be suboptimal since real FRBs do not always cover the full frequency range.
Next, we introduce a ``sub-band'' search for searching a configurable set of subbands.

\subsection{Parameterization of subbands}

Our dedispersion config files contain a length-$(R+1)$ integer-valued array of subband counts $C_l$.\footnote{In 
the code, $R$ is denoted {\tt pf\_rank}, $l$ is denoted {\tt pf\_level}, and $C_l$ is denoted {\tt frequency\_subband\_counts}.}
Here $0 \le R \le r$, where $r$ is the tree rank.
The counts $C_l$ parameterize a set of subbands as follows:
\begin{itemize}
\item Frequency subbands are indexed by a pair $(l,s)$ where $0 \le l \le R$ and $0 \le s < C_l$.
\item Level $l=0$ is special: subband $(0,s)$ corresponds to tree-freq range
\begin{equation}
2^{r-R} s \le \mbox{(tree-freq index)} < 2^{r-R} (s+1)  \label{eq:subband_l0}
\end{equation}
\item At levels $l>0$, subband $(l,s)$ corresponds to tree-freq range
\begin{equation}
2^{r-R+l-1} s \le \mbox{(tree-freq index)} < 2^{r-R+l-1} (s+2)  \label{eq:subband_l1}
\end{equation}
Thus, level 0 consists of non-overlapping bands whose fractional width is $1/2^R$.
Levels $l>0$ consist of bands whose fractional width is $1/2^{R-l}$, and overlap by a factor 2.
\item
To avoid ``overflowing'' the band, $C_l$ must satisfy
\begin{equation}
0 \le C_l \le \begin{cases}
2^R & \mbox{for } l=0 \\
2^{R-l+1}-1 & \mbox{for } l>0
\end{cases}
\end{equation}
A level with $C_l=0$ is allowed, and simply contributes no subbands.
We require $C_R=1$, i.e.~we always search the full frequency band.
The case $C_l=\{1\}$ (with $R=0$) corresponds to disabling subbands, and only searching the full band.
\item The details of this scheme, including the ``specialness'' of level $l=0$, are
dictated by convenience in the GPU kernel.
\end{itemize}

\subsection{Subbanded dedispersion}

Subbanded dedispersion is an operation whose input array is the same as ordinary (non-subbanded) tree 
dedispersion (\ref{sec:tree_dedispersion}), i.e.\ a shape $(2^r, N_{\rm time})$ array where the first index is tree-freq.
The output array has shape $(2^{r-R}, M, N_{\rm time})$, where $M$ is the number of ``multiplets'':
\begin{equation}
M \equiv \sum_{l=0}^R 2^l C_l
\end{equation}
Each subband $(l,s)$ ``owns'' a shape $(2^{r-R}, 2^l, N_{\rm time})$ subarray of the output.
The first two axes correspond to $2^{r-R+l}$ dispersion delays across the subband, split into a ``coarse'' index
$0 \le d_{\rm hi} < 2^{r-R}$ and a ``fine'' index  $0 \le d_{\rm lo} < 2^l$.
(This index-splitting ensures that the combined output of all subbands is a regular $(2^{r-R}, M, N_{\rm time})$ array,
rather than a ``ragged'' object.)

Conceptually, subband dedispersion works as follows.
For each subband $(l,s)$, we apply tree dedispersion to the appropriate tree-freq range (either Eq.\ (\ref{eq:subband_l0}) 
or (\ref{eq:subband_l1})). We also apply a time lag so that the time index corresponds to the extrapolated arrival time at 
the highest tree-freq in the full band, not the highest tree-freq in the subband.
For an efficient implementation, we run tree dedispersion as usual (Eq.\ \ref{eq:DD_composition}), and extract per-subband
dedispersion outputs from intermediate arrays ``on the fly'', rather than dedispersing each subband ``from scratch''.

Formally, the details are as follows:

\begin{itemize}

\item {\bf Case 1: $l=0$ or even $s$.}
In this case, the subband corresponds to an ``aligned'' tree-freq range of the form:
\begin{equation}
2^{r-R+l} f \le \mbox{(tree-freq index)} < 2^{r-R+l}(f+1) \hspace{1cm} \mbox{(where $0 \le f < 2^{R-l}$)}
\end{equation}
where the coarse-freq index $f$ is given by:
\begin{equation}
f = \begin{cases} s & \mbox{if } l=0 \\ s/2 & \mbox{if } l > 0 \end{cases}
\end{equation}
The subband owns a shape $(2^{r-R}, 2^l, N_{\rm time})$ subarray of the output (denote by \verb|out|).

In this case, the per-subband dedispersion outputs already arise as intermediates in tree dedispersion
(Eq.\ \ref{eq:DD_composition}); we just need to apply a time lag.
Let \verb|tmp| be the output array of $\DD(k-1)$, where $k=r-R+l$. 
Then \verb|tmp| has shape $(2^{R-l}, 2^{r-R+l}, N_{\rm time})$.
To populate the \verb|out| array, we slice/reshape \verb|tmp| and apply time lags:
\begin{equation}
{\tt out}\big[ d_{\rm hi}, d_{\rm lo}, t \big] = {\tt tmp}\big[ f, d, t - T \big]
\end{equation}
where on the RHS we define:
\begin{equation}
d \equiv d_{\rm hi} 2^l + d_{\rm lo}   \hspace{1.5cm} T \equiv d (2^{R-l}-1-f)
\end{equation}

\item {\bf Case 2: $l>0$ and odd $s$.}
In this case, the subband corresponds to a ``half-aligned'' tree-freq range of the form:
\begin{equation}
2^{r-R+l-1} (2f+1) \le \mbox{(tree-freq index)} < 2^{r-R+l-1}(2f+3) \hspace{1cm} \mbox{(where $0 \le f < 2^{R-l}-1$)}
\end{equation}
where the coarse-freq index $f$ is given by $f \equiv (s-1)/2$.
The subband owns a shape $(2^{r-R}, 2^l, N_{\rm time})$ subarray of the output (denote by \verb|out|).

This case is a little more complicated, since the per-subband dedispersion outputs are ``one DD-step away''
from the intermediate arrays in tree dedispersion.
Let \verb|tmp| be the output array of $\DD(k-1)$, where $k=r-R+l-1$. 
Then \verb|tmp| has shape $(2^{R-l+1}, 2^{r-R+l-1}, N_{\rm time})$.
\begin{align}
{\tt out}\big[ d_{\rm hi}, 2 d_{\rm lo}, t \big] 
  &= \frac{1}{\sqrt{2}} \Big( {\tt tmp}\big[ 2f+1, d, t - T_0 - d \big] + {\tt tmp}\big[ 2f+2, d, t - T_0 \big] \Big) \nn \\
{\tt out}\big[ d_{\rm hi}, 2 d_{\rm lo} + 1, t \big] 
  &= \frac{1}{\sqrt{2}} \Big( {\tt tmp}\big[ 2f+1, d, t - T_1 - d - 1 \big] + {\tt tmp}\big[ 2f+2, d, t - T_1 \big] \Big) 
\end{align}
where on the RHS we define:
\begin{equation}
d \equiv d_{\rm hi} 2^{l-1} + d_{\rm lo}
  \hspace{1.5cm} T_0 = d (2^{R-l+1} - 2f - 3) 
  \hspace{1.5cm} T_1 = T_0 + (2^{R-l} - f - 2)
\end{equation}
(Warning: the quantities $f,k,d_{\rm lo}$ have slightly different meanings in cases 1 and 2!)

\end{itemize}
Note that subband dedispersion includes the full band (via $C_R=1$), but the full-band output array is
``reshaped'' to $(2^{r-R}, 2^R, N_{\rm time})$, whereas in the previous section it had shape $(2^r, N_{\rm time})$.

\end{document}
